Investigation on Parallel Training of Deep Neural Network using Model Averaging

### 2015-06-25 paper.tex
remove the butterfly mixing

% Example definitions.
% --------------------
\def\x{{\mathbf x}}
\def\L{{\cal L}}

% Title.
% ------
\title{Investigation on Parallel Training of Deep Neural Network using Model Averaging}
%
% Single address.
% ---------------
\name{Hang Su$^{1,2}$, Haoyu Chen$^1$, Haihua Xu$^3$}
\address{$^1$ International Computer Science Institute, Berkeley, California, US \\
$^2$ Dept. of Electrical Engineering \& Computer Science, University of California, Berkeley, CA, USA \\
$^3$ Nanyang Technological University, Singapore \\
{\small \tt \{suhang3240@gmail.com\}}
}
%
% For example:
% ------------
%\address{School\\
%	Department\\
%	Address}
%
% Two addresses (uncomment and modify for two-address case).
% ----------------------------------------------------------
%\twoauthors
%  {A. Author-one, B. Author-two\sthanks{Thanks to XYZ agency for funding.}}
%	{School A-B\\
%	Department A-B\\
%	Address A-B}
%  {C. Author-three, D. Author-four\sthanks{The fourth author performed the work
%	while at ...}}
%	{School C-D\\
%	Department C-D\\
%	Address C-D}
%
\begin{document}
%\ninept
%
\maketitle
%
\begin{abstract}
In this paper we apply model averaging to parallel training of deep neural network (DNN). 
Parallelization is done in a model averaging manner. Data is partitioned and distributed to different nodes for 
local model updates, and model averaging across nodes is done every few minibatches.

We use multiple GPUs for data parallelization, and Message Passing Interface (MPI) for communication between nodes. 
We investigate the effectiveness of Natrual Gradient for Stochasitc Gradient Descent (NG-SGD) in the model averaging
framework, and compare two learning rate schedule strategies for model update. It is shown that NG-SGD leads to better
convergence in model averaging framework. On the 300h Swithboard dataset, a 10 times speedup is achieved using 16 GPUs
and 17 times speedup using 32 GPUs with a small decoding accuracy loss -- 0.4\% degradation in word error rate 
(WER) respectively. We also introduce a butterfly mixing strategy that can further save communication costs in model 
averaging framework.
\end{abstract}
%
\begin{keywords}
Parallel training, model averaging, deep neural network, butterfly mixing
\end{keywords}
%
\section{Introduction}
\label{sec:intro}
Deep Neural Networks (DNN) has shown its effeciveness in several machine learning tasks, espencially in speech
recognition. The large model size and massive training examples make DNN a powerful model for classification. However,
these two factors also slow down the training procedure of DNNs.

@@

Our approach is mainly based on the NG-SGD with model averaging. We utilize multiple GPUs in neural networks training via 
MPI, which allows us to perform model averaging more frequently and efficiently. Currently, this work simplifies the model 
training procedure in the following aspects:
1. It does not add a warm-up phase where only single thread is used for model update \cite{seide20141,povey2014parallel};
2. It does not tune minibatch size to utilize parallelization power;

On the other hand, it requires:
1. A proper implementation of NG-SGD, as is introduced in \cite{povey2014parallel};
2. A bigger initial learning rate and a proper model averaging frequency;

In Section 2 we introduce the model averaging approach. In Section 3 we introduce butterfly-mixing strategie. Section 4 
explains NG-SGD. Section 5 records experimental results on different setups and Section 6 concludes.

@@

On the other hand, if we choose to average the parameters rather than gradients, it is not necessary to exchange data that often. 
Currently, there is no straight forward theory that guarantees convergence, but we would like to explore a bit why this strategy 
shoule work, just as we observe in the experiments.

First, in the extreme case where model parameters are averaged after each minibatch update, model averaging is equivalent to 
gradient averaging. If we view the difference of model parameters as gradient in a special space.

@@

\section{Model Averaging Strategy}
\subsection{All-reduce}
All-reduce strategy collects the weights from all the nodes. The communication is bounded by 
the bandwidth. Fig~\ref{fig:allreduce} is an example of all-reduce with 4 nodes. One node is responsible for
collecting all the model parameters. It will average the weights and then broadcast it to all the nodes. 
This approach compute the average weight based on all the information in computing nodes. Thus, we expect 
the converge speed to be the fastest.
\begin{figure}[htb]
  \centering
  \includegraphics[width=0.5\textwidth]{allreduce.png}
  \caption{All-reduce network}
  \label{fig:allreduce}
\end{figure}

This operation could be easily implemented by MPI\_Allreduce

\subsection{Butterfly mixing}
Butterfly mixing was proposed in \cite{zhao2013butterfly} to interleave communication with computation. In each
reduce period, information (gradient / model) is exchanged between each node and its friend node. The pair of nodes
are shuffled after every reduce so that information in each node can propagate to all other nodes in $log(numNodes)$ periods.

Butterfly mixing is a reduction strategy proposed in \cite{zhao2013butterfly}. It reduces the communication load in 
every iteration: one node would only send and receive message from one other node. The pairing of nodes are shuffled after
each reduction so that information collected in each nodes will spread to all other nodes. Fig~\ref{fig:butterfly} 
is an example of butterfly mixing with 4 nodes. Butterfly mixing requires much less communication than all-reduce 
strategy. However, each node in the network only collects information from two computing nodes at a time.
It takes $log(numNodes)$ communication to spread the message to the whole network. Thus, the converge speed 
of butterfly mixing would be slower than all-reduce strategy. 
\begin{figure}[htb]
  \centering
  \includegraphics[width=0.5\textwidth]{butterfly.png}
  \caption{Butterfly mixing network}
  \label{fig:butterfly}
\end{figure}

MPI implementation

\subsection{Data exchange analysis}

\section{Natural Gradient for Model Update}
In stochastic gradient descent (SGD), the learning rate is often assumed to be a scalar $\alpha_t$ that may change over time,
the update formula for model parameters $\theta_{t}$ is
\begin{equation}
\theta_{t+1} = \theta_{t} + \alpha_t g_t
\end{equation}
where $g_t$ is the gradient.

@@

For large scale speech recognition, it is impossible to estimate Fisher information matrix and perform inversion, 
so it is necessary to approximate the inverse Fisher information matrix directly. We do not include too much 
detail here because this is not the main focus of this report. Details about implementation of NG-SGD could 
be found in \cite{povey2014parallel}.

\section{Experimental Results}
\subsection{Setup}
In this paper, we report speech recognition results on the 300 hour Switchboard conversational telephone speech task 
(Switchboard-1 Release 2). We use MSU-ISIP release of the Switchboard segmentations and transcriptions (date 11/26/02),
together with the Mississippi State transcripts2 and the 30Kword lexicon released with those transcripts. 
The lexicon contains pronunciations for all words and word fragments in the training data. We use the Hub5 ’00 data for
evaluation. Specifically, we use the  the development set and Hub5 ’01 (LDC2002S13) data as a separate test set.

@@

Work in this paper is built on top of the Kaldi nnet1 setup and the NG-SGD method introduced in nnet2 setup. 
Details of DNN training follows Section 2.2 in \cite{vesely2013sequence}. In this paper, we use 6 hidden layers, where each 
hidden layer has 2048 neurons with sigmoids. Input layer is 440 dimension (i.e. the context of 11 fMLLR frames), 
and output layer is 8806 dimension. Mini-batch SGD is used for backpropagation and the minibatch is set to 1024 for all
the experiments. By defult, DNNs are initialized with stacked restricted Boltzmann machines (RBMs) that are pretrained 
in a greedy layerwise fashion \cite{hinton2006fast}. Comparison between random initialization and RBM-initialization 
in model averaging framework is reported in Section~\ref{sec:init}.

@@

\subsection{Averaging frequency}
Due to the limitation of computing resource, we did not tune averaging frequency in our setup. It is set to 1 averaging
per 20 minibatches by default. We would expect a tradeoff between lower averaging frequency (i.e. better speedup) and
better training accuracy. But the optimal point is yet to be explored.

@@

\subsection{Initialization Matters}
\label{sec:init}
Compare random init and dbn init, single and parallell.

@@

\subsection{Speedup Comparison}
scaling factor
\begin{figure}[htb]
  \centering
  \includegraphics[width=0.5\textwidth]{scaling.png}
  \caption{Scaling factor and speedup factor v.s. number of gpus}
  \label{fig:scaling}
\end{figure}

@@

We acknowledge the Texas Advanced Computing Center (TACC) at The University of Texas at Austin for providing 
HPC resources that have contributed to the research results reported within this paper (URL: http://www.tacc.utexas.edu).

>>>>

% Example definitions.
% --------------------
\def\x{{\mathbf x}}
\def\L{{\cal L}}
\newcommand{\overbar}[1]{\mkern 1.5mu\overline{\mkern-1.5mu#1\mkern-1.5mu}\mkern 1.5mu}

% Title.
% ------
\title{Investigation on Parallel Training of Deep Neural Network using Model Averaging}
%
% Single address.
% ---------------
\name{Hang Su$^{1,2}$, Haoyu Chen$^1$, Haihua Xu$^3$}
\address{$^1$ International Computer Science Institute, Berkeley, California, US \\
$^2$ Dept. of Electrical Engineering \& Computer Science, University of California, Berkeley, CA, USA \\
$^3$ Nanyang Technological University, Singapore \\
{\small \tt \{suhang3240@gmail.com\}}
}
%
% For example:
% ------------
%\address{School\\
%	Department\\
%	Address}
%
% Two addresses (uncomment and modify for two-address case).
% ----------------------------------------------------------
%\twoauthors
%  {A. Author-one, B. Author-two\sthanks{Thanks to XYZ agency for funding.}}
%	{School A-B\\
%	Department A-B\\
%	Address A-B}
%  {C. Author-three, D. Author-four\sthanks{The fourth author performed the work
%	while at ...}}
%	{School C-D\\
%	Department C-D\\
%	Address C-D}
%
\begin{document}
%\ninept
%
\maketitle
%
\begin{abstract}
In this work we apply model averaging to parallel training of deep neural network (DNN). 
Parallelization is done in a model averaging manner. Data is partitioned and distributed to different nodes for 
local model updates, and model averaging across nodes is done every few minibatches.

We use multiple GPUs for data parallelization, and Message Passing Interface (MPI) for communication between nodes,
which allows us to perform model averaging frequently without losing much time on communication.
We investigate the effectiveness of Natrual Gradient for Stochasitc Gradient Descent (NG-SGD) in the model averaging
framework, and compare two learning rate schedule strategies for model update. It is shown that NG-SGD leads to better
convergence in model averaging framework. On the 300h Swithboard dataset, a 9.3 times speedup is achieved using 16 GPUs
and 17 times speedup using 32 GPUs with limited decoding accuracy loss. 
\end{abstract}
%
\begin{keywords}
Parallel training, model averaging, deep neural network
\end{keywords}
%
\section{Introduction}
\label{sec:intro}
Deep Neural Networks (DNN) has shown its effeciveness in several machine learning tasks, espencially in speech
recognition. The large model size and massive training examples make DNN a powerful model for classification. However,
these two factors also slow down the training procedure of DNNs.

@@

Our approach is mainly based on the NG-SGD with model averaging. We utilize multiple GPUs in neural networks training via 
MPI, which allows us to perform model averaging more frequently and efficiently. Unlike the other approach \cite{seide20141},
we do not use a warm-up phase where only single thread is used for model update. (Admittedly, this might lead to further improvement). 
In this work, we conduct a lot of experiments and compare different setups in model averaging framework.

In Section 2, we introduce related works on NG-SGD. Section 3 describe the model averaging approach and some 
intuition on the analysis. Section 4 records experimental results on different setups and Section 5 concludes.

@@

On the other hand, if we choose to average the parameters rather than gradients, it is not necessary to exchange data that often. 
Currently, there is no straight forward theory that guarantees convergence, but we would like to explore a bit why this strategy 
should work, just as we observe in the experiments.

First, in the extreme case where model parameters are averaged after each weight update, model averaging is equivalent to 
gradient averaging. Furthermore, if model averaging is done every $n$ minibatch based weight update, model update formula could
be written as
\begin{equation}
\begin{split}
\theta_{t+n} &= \theta_{t} +\sum_{i=0}^{n-1} \alpha g_{t+i}\\
&= \theta_{t} +\sum_{i=0}^{n-1} \alpha \frac{\partial}{\partial\theta} F(x;\theta_{t+i})
\end{split}
\end{equation}
\begin{equation}
\overbar{\theta_{t+n}} = \overbar{\theta_{t}} +\sum_{i=0}^{n-1} \alpha \frac{\partial}{\partial\theta} \overbar{F}(x;\theta_{t+i})
\end{equation}
where $\theta$ is the model parameter and $\alpha$ is learning rate. If changes in model parameter $\theta$ is limited 
within $n$ updates, this approach could be seen as an approximation to gradient averaging.

@@

Fig~\ref{fig:allreduce} is an example of all-reduce with 4 nodes. This operation could be easily implemented by MPI\_Allreduce.
\begin{figure}[htb]
  \centering
  \includegraphics[width=0.5\textwidth]{allreduce.png}
  \caption{All-reduce network}
  \label{fig:allreduce}
\end{figure}

\section{Natural Gradient for Model Update}
This section introduces the idea proposed in \cite{povey2014parallel}.

In stochastic gradient descent (SGD), the learning rate is often assumed to be a scalar $\alpha_t$ that may change over time,
the update formula for model parameters $\theta_{t}$ is
\begin{equation}
\theta_{t+1} = \theta_{t} + \alpha_t g_t
\end{equation}
where $g_t$ is the gradient.

@@

For large scale speech recognition, it is impossible to estimate Fisher information matrix and perform inversion, 
so it is necessary to approximate the inverse Fisher information matrix directly. Details about the theory and 
implementation of NG-SGD could be found in \cite{povey2014parallel}.

\section{Experimental Results}
\subsection{Setup}
In this work, we report speech recognition results on the 300 hour Switchboard conversational telephone speech task 
(Switchboard-1 Release 2). We use MSU-ISIP release of the Switchboard segmentations and transcriptions (date 11/26/02),
together with the Mississippi State transcripts2 and the 30Kword lexicon released with those transcripts. 
The lexicon contains pronunciations for all words and word fragments in the training data. We use the Hub5 ’00 data for
evaluation. Specifically, we use the  the development set and Hub5 ’01 (LDC2002S13) data as a separate test set.

@@

Work in this paper is built on top of the Kaldi nnet1 setup and the NG-SGD method introduced in nnet2 setup. 
Details of DNN training follows Section 2.2 in \cite{vesely2013sequence}. In this work, we use 6 hidden layers, where each 
hidden layer has 2048 neurons with sigmoids. Input layer is 440 dimension (i.e. the context of 11 fMLLR frames), 
and output layer is 8806 dimension. Mini-batch SGD is used for backpropagation and the minibatch is set to 1024 for all
the experiments. By defult, DNNs are initialized with stacked restricted Boltzmann machines (RBMs) that are pretrained 
in a greedy layerwise fashion \cite{hinton2006fast}. Comparison between random initialization and RBM-initialization 
in model averaging framework is reported in Section~\ref{sec:init}.

@@

\subsection{Switchboard Results}
scaling factor
\begin{figure}[htb]
  \centering
  \includegraphics[width=0.5\textwidth]{scaling.png}
  \caption{Scaling factor and speedup factor v.s. number of gpus}
  \label{fig:scaling}
\end{figure}

\subsection{Initialization Matters}
\label{sec:init}
Compare random init and dbn init.

\subsection{Averaging frequency}
Averaging frequency here is defined as the number of minibatch-SGD performed per model averaging.
Due to the limitation of computing resource, we only did preliminary experiments on this. Minibatch size of 1024 
is set as default, and we compare averaging frequency of 10 and 20. It is shown that an averaging frequency of 10 gives
slight worse speedup but a better decoding WER. The tradeoff between lower averaging frequency (i.e. better speedup) and
better training accuracy is within expectation in that frequent model averaging means steady gradient estimation.

\begin{table}
  \centering
  \begin{tabular}{c|c|c|c}
    \hline
    frequency   & Speedup   & SWB   & CallHome \\
    \hline
    baseline    &   --      & 14.7  & 26.8\\
    \hline
    10          & 9.32      & 15.1  & 27.0 \\
    \hline
    20          & 10.07     & 15.8  & 28.0 \\
    \hline
  \end{tabular}
  \caption{Comparing different averaging frequencies}
  \label{tab:init}
\end{table}

\subsection{Minibatch Size}
\begin{table}
  \centering
  \begin{tabular}{c|c|c|c|c|c}
    \hline
    & \multicolumn{2}{c|}{SWB}    & \multicolumn{2}{c}{CallHome} & Speedup\\
    \hline
    nodes       &  1    &   16    &    1       &   16   &    \\
    \hline
    256         &  --      & 15.1  & 27.0     &   & \\
    \hline
    1024        & 9.32     & 15.1  & 27.0     &   & \\
    \hline
  \end{tabular}
  \caption{Comparing different minibatch size}
  \label{tab:init}
\end{table}

@@

@@

@@

We acknowledge the Texas Advanced Computing Center (TACC) at The University of Texas at Austin for providing 
HPC resources that have contributed to the research results reported within this work (URL: http://www.tacc.utexas.edu).
